---
author: "MPI"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
title: Acceptable Inputs
vignette: >
  %\VignetteIndexEntry{Acceptable Inputs}
  %\VignetteEngine{knitr::knitr}
editor_options:
  chunk_output_type: console
---

# Acceptable Inputs

## Run cpt with a trackframe

check as.trackframe method for data.frame
```{r eval=FALSE, include=TRUE}
?as.trackframe.data.frame
```


Transform data to a trackframe. Trackframe always use x/y coordinates as cartesian coordinates with easting and northing.
```{r}
library(trackframe)
projected_crs <- "EPSG:32632"
tf <- as.trackframe(path_matrix, crs = projected_crs)
class(tf)
```


```{r}
library(cpt)
set.seed(2025L)
cpt_tf <- change_point_test(tf, alpha = 0.05, q = 3, n = 500, min_move_dist = 0)
class(cpt_tf)
```

Extract the change points into a summarized format
```{r}
summary(cpt_tf)
```


## Run cpt with a trackframe and multiple IDs

Generate 3 tracks with different sizes
```{r}
data("paths_trackframe", package = "trackframe")
class(paths_trackframe)
```

```{r}
set.seed(2025L)
cpt_mt <- change_point_test(paths_trackframe, alpha = 0.05, q = 3, N = 500, min_move_dist = 0)
class(cpt_mt)
```

Extract the change points into a summarized format
```{r}
summary(cpt_mt)
```

```{r}
head(select_id(cpt_mt, "track_1"))
tail(select_id(cpt_mt, "track_3"))
```


## Run cpt with xyt data.frame

Detect change points
```{r}
data("path_data_frame", package = "trackframe")
class(path_data_frame)
set.seed(2025L)
cpt <- change_point_test(path_data_frame, alpha = 0.05, q = 3, N = 500)
class(cpt)
```

Alternatively, in order to remove the warning coerce to trackframe first by setting crs.

```{r}
projected_crs <- "EPSG:32632"
tf <- as.trackframe(path_matrix, crs = projected_crs)
set.seed(2025L)
cpt_tf <- change_point_test(tf, alpha = 0.05, q = 3, N = 500)
all.equal(cpt$cp_id, cpt_tf$cp_id)
```

Extract the change points into a summarized format
```{r}
summary(cpt)
```

## Run cpt with move2

Detect change points
```{r}
data("path_move2", package = "trackframe")
class(path_move2)
set.seed(2025L)
path_move2 <- sf::st_transform(path_move2, projected_crs)
class(path_move2) <- c("move2", "sf", "data.frame")
cpt <- change_point_test(path_move2, alpha = 0.05, q = 3, N = 100)
class(cpt)
```

Extract the change points into a summarized format
```{r}
tail(summary(cpt), n = 5)
```

## Run cpt with sftrack

Detect change points
```{r}
data("path_sftrack", package = "trackframe")
class(path_sftrack)
path_sftrack <- sf::st_transform(path_sftrack, projected_crs)
class(path_sftrack) <- c("sftrack", "sf", "data.frame")
set.seed(2025L)
cpt <- change_point_test(path_sftrack, alpha = 0.05, q = 3, N = 500)
class(cpt)
```

Extract the change points into a summarized format
```{r}
summary(cpt)
```



## Run cpt with cocomo format

Create cocomo format data
```{r}
padding <- 25
n_times <- 20 + 3 * padding
n <- 2

xs <- matrix(
  c(
    rep(-1, padding),
    c(-1, -0.9, -0.8, -0.7, -0.6, -0.5, -0.4, -0.3, -0.2, -0.1),
    rep(0, padding),
    c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1),
    rep(1, padding),
    rep(0, n_times)
  ),
  n, n_times, byrow = TRUE
)
ys <- matrix(
  rep(0, n * n_times),
  n, n_times, byrow = TRUE
)
timestamps <- as.POSIXct(1:n_times)

ids <- rbind.data.frame(
  list("id_code" = 'VCVM001', "age" = 10, "sex" = "m"),
  list("id_code" = 'WRTH', "age" = 5, "sex" = "f")
)
```

Transform to trackframe
```{r}
cocomo_tf <- cocomo_as_tf(xs, ys, timestamps, ids)
head(cocomo_tf)
class(cocomo_tf)
```

Detect change points
```{r}
set.seed(2025L)
cpt <- change_point_test(cocomo_tf, alpha = 0.05, q = 3, N = 500)
class(cpt)
```

Extract the change points into a summarized format (no change points found)
```{r}
summary(cpt)
```
