---
author: "MPI"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette cpt package}
  %\VignetteEngine{knitr::knitr}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300)
```


# Introduction

## Install cpt package

```{r eval=FALSE, include=TRUE}
library("remotes")
install_github('livingingroups/travelpaths-devel/pkgs/trackframe')
install_github('livingingroups/travelpaths-devel/pkgs/cpt')
```


## Load packages

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(changepointpermtest2009)
library(trackframe)
```


**cpt** package currently supports xyt matrices (a matrix or data frame with columns for x-coordinates, y-coordinates, and timestamps) and trackframes (see pkg trackframe) as input formats.

## Run cpt with xyt

Detect change points
```{r}
data("path_trackframe", package = "trackframe")
class(path_trackframe)
set.seed(2025L)
cpt <- change_point_test_xyt(
  easting = path_trackframe[, "easting"],
  northing = path_trackframe[, "northing"],
  time = path_matrix[, "time"],
  alpha = 0.05,
  q = 3,
  N = 500
)
class(cpt)
```

Extract the change points into a summarized format
```{r}
summary(cpt)
```


### Plot the change points

TODO





## Extract pvalues from Permutation Test

Get probability values instead of binary indicators
```{r}
data("cpttestdata_tiny")
pvalues <- change_point_test_pvalue(cpttestdata_tiny, q_max = 5, N = 500)
tail(pvalues)
```
