---
author: "MPI"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette cpt package}
  %\VignetteEngine{knitr::knitr}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300)
```


# Introduction

## Install cpt package

```{r eval=FALSE, include=TRUE}
library("remotes")
install_github('livingingroups/travelpaths-devel/pkgs/trackframe')
install_github('livingingroups/travelpaths-devel/pkgs/cpt')
```


## Load packages

```{r echo=TRUE, message=FALSE, warning=FALSE}
library("cpt")
library("trackframe")
# library(ggplot2)
```


**cpt** package currently supports xyt matrices (a matrix or data frame with columns for x-coordinates, y-coordinates, and timestamps) and trackframes (see pkg trackframe) as input formats.

## Run cpt with xyt

Detect change points
```{r}
data("path_trackframe", package = "trackframe")
class(path_trackframe)
set.seed(2025L)
cpt <- change_point_test_xyt(
  easting = path_trackframe[, "easting"],
  northing = path_trackframe[, "northing"],
  time = path_matrix[, "time"],
  alpha = 0.05,
  q = 3,
  n = 500
)
class(cpt)
```

Extract the change points into a summarized format
```{r}
summary(cpt)
```


### Plot the change points

TODO


## Run cpt with a trackframe

check as.trackframe method for data.frame
```{r eval=FALSE, include=TRUE}
?as.trackframe.data.frame
```


Transform data to a trackframe. Trackframe always use x/y coordinates as cartesian coordinates with easting and northing. 
```{r}
tf <- as.trackframe(path_matrix, crs_input = 4326)
class(tf)
```


```{r}
set.seed(2025L)
cpt_tf <- change_point_test(tf, alpha = 0.05, q = 3, n = 500, tol = 0)
class(cpt_tf)
```

Extract the change points into a summarized format
```{r}
summary(cpt_tf)
```


## Run cpt with a trackframe and multiple IDs

Generate 3 tracks with different sizes
```{r}
data("paths_trackframe", package = "trackframe")
class(paths_trackframe)
```

```{r}
set.seed(2025L)
cpt_mt <- change_point_test(paths_trackframe, alpha = 0.05, q = 3, n = 500, tol = 0)
class(cpt_mt)
```

Extract the change points into a summarized format
```{r}
summary(cpt_mt)
```

```{r}
head(select_id(cpt_mt, "track_1"))
tail(select_id(cpt_mt, "track_3"))
```


## Run cpt with xyt data.frame

Detect change points
```{r}
data("path_data_frame", package = "trackframe")
class(path_data_frame)
set.seed(2025L)
cpt <- change_point_test(path_data_frame, alpha = 0.05, q = 3, n = 500)
class(cpt)
```

Alternatively, in order to remove the warning coerce to trackframe first by setting crs.

```{r}
tf <- as.trackframe(path_data_frame, crs_input = 4326)
set.seed(2025L)
cpt_tf <- change_point_test(tf, alpha = 0.05, q = 3, n = 500)
all.equal(cpt$sig, cpt_tf$sig)
```

Extract the change points into a summarized format
```{r}
summary(cpt)
```

## Run cpt with move2

Detect change points
```{r}
data("path_move2", package = "trackframe")
class(path_move2)
set.seed(2025L)
cpt <- change_point_test(path_move2, alpha = 0.05, q = 3, n = 100)
class(cpt)
```

Extract the change points into a summarized format
```{r}
tail(summary(cpt), n = 5)
```

## Run cpt with sftrack

Detect change points
```{r}
data("path_sftrack", package = "trackframe")
class(path_sftrack)
set.seed(2025L)
cpt <- change_point_test(path_sftrack, alpha = 0.05, q = 3, n = 500)
class(cpt)
```

Extract the change points into a summarized format
```{r}
summary(cpt)
```



## Run cpt with cocomo format

Create cocomo format data
```{r}
padding <- 25
n_times <- 20 + 3 * padding
n <- 2

xs <- matrix(
  c(
    rep(-1, padding),
    c(-1, -0.9, -0.8, -0.7, -0.6, -0.5, -0.4, -0.3, -0.2, -0.1),
    rep(0, padding),
    c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1),
    rep(1, padding),
    rep(0, n_times)
  ),
  n,
  n_times,
  byrow = TRUE
)
ys <- matrix(
  rep(0, n * n_times),
  n,
  n_times,
  byrow = TRUE
)
timestamps <- as.POSIXct(1:n_times)

ids <- rbind.data.frame(
  list("id_code" = 'VCVM001', "age" = 10, "sex" = "m"),
  list("id_code" = 'WRTH', "age" = 5, "sex" = "f")
)
```

Transform to trackframe
```{r}
cocomo_tf <- cocomo_as_tf(xs, ys, timestamps, ids)
head(cocomo_tf)
class(cocomo_tf)
```

Detect change points
```{r}
set.seed(2025L)
cpt <- change_point_test(cocomo_tf, alpha = 0.05, q = 3, n = 500)
class(cpt)
```

Extract the change points into a summarized format (no change points found)
```{r}
summary(cpt)
```




## Extract pvalues from Permutation Test

Get probability values instead of binary indicators
```{r}
data("cpttestdata")
pvalues <- change_point_test_pvalue(cpttestdata, q_max = 5, n = 500)
tail(pvalues)
```

# Recreate Analysis as in TODO cite paper

Load data
```{r}
data("cptfiguredata_tf")
str(cptfiguredata_tf)
tf <- cptfiguredata_tf
#FIXME
tf$y <- tf$y * (-1)
```


```{r}
library(ggplot2)
```

### Fig  3

```{r, out.width="100%", fig.width=7}
tf3 <- tf[startsWith(tf$track_id, '3'), ]

ggplot() +
  geom_path(
    data = tf3[order(time(tf3)), ],
    mapping = aes(x = easting(tf3), y = northing(tf3), group = track_id)
  ) +
  coord_fixed()
# TODO:
# - show starting point and direction.
# - facet wrap instead of relying on fake coords being non-overlapping
```

### Fig 4


```{r}
tf4 <- tf[startsWith(tf$track_id, '4'), ]
# b and d seem to match up well
# a and c seem to be missing some changepoints
cp <- change_point_test(
  tf4,
  alpha = .01,
  n = 10000,
  q = 6
)
```

```{r, out.width="100%"}

ggplot() +
  geom_path(
    data = tf4[order(time(tf4)), ],
    mapping = aes(x = easting(tf4), y = northing(tf4), group = track_id)
  ) +
  coord_fixed() +
  geom_point(data = cp[cp$sig == 1, ], aes(x = x, y = y), color = "red")
```

### Fig 5

```{r, out.width="100%", fig.height=10, , fig.width=5}
n <- 100
tf_4a <- tf[tf$track_id == '4a', ]
set.seed(2025)
P = change_point_test_pvalue(
  # P = cpt:::change_point_test_pvalue_old( #gives different result due to different calculation order
  tf_4a,
  q_max = 10,
  n = n
)

# plot results (- log(P) vs k for q = 1, .. , q_max)

no.of.t <- NROW(P)
k <- c(1:no.of.t)
s <- seq(no.of.t - 1)

par(mfrow = c(5, 2))
for (i in seq_len(NCOL(P))) {
  plot(
    k,
    -log(P[k, i]),
    xlim = c(0, no.of.t + 1),
    ylim = c(0, log(n)),
    pch = 4,
    main = sprintf("q = %i", i)
  )
  y <- -log(P[k, i])
  segments(k[s], y[s], k[s + 1], y[s + 1])
  abline(h = -log(0.10), col = "blue")
  abline(h = -log(0.05), col = "red")
  abline(h = -log(0.01), col = "green")
}
```


### Fig 6 

```{r}
# Pretty different
calculate_n_cp_by_q <- function(tf, q = seq_len(10), alpha = 0.01) {
  # tf <- tf[startsWith(tf$track_id, '4'), ]
  if (length(attr(tf, 'id')) != 1) {
    stop('only implemented for 1 id col')
  }
  do.call(
    rbind,
    lapply(seq_len(10), function(q) {
      cp <- change_point_test(
        tf,
        alpha = alpha,
        q = q
      )
      # cp <- do.call(rbind, lapply(seq_len(cp), function(i) cbind(cp, "track_id" = i)))
      df <- as.data.frame(rowsum(cp$sig, as.factor(cp[, attr(tf, 'id')])))
      names(df) <- 'n_cp'
      df$track_id <- rownames(df)
      rownames(df) <- NULL
      df$q <- q
      df
    })
  )
}

fig6_data <- calculate_n_cp_by_q(tf4)
```

```{r, out.width="100%"}
ggplot(data = fig6_data, mapping = aes(x = q, y = n_cp, group = track_id, colour = track_id)) +
  geom_smooth(span = .4) +
  geom_point()
```

### Fig 7

```{r}
tf7 <- tf[startsWith(tf$track_id, '7'), ]
cp <- change_point_test(
  tf7,
  alpha = .05,
  q = 4
)
```

```{r, out.width="100%"}
ggplot() +
  geom_path(
    data = tf7[order(time(tf7)), ],
    mapping = aes(x = easting(tf7), y = northing(tf7), group = track_id)
  ) +
  coord_fixed() +
  geom_point(data = cp[cp$sig == 1, ], aes(x = x, y = y), color = "red")
```

### Fig 8

```{r, out.width="100%", fig.height=10, , fig.width=5}
n <- 100
tf_7a <- tf[tf$track_id == '7a', ]
set.seed(2025)
data <- tf_7a
P = change_point_test_pvalue(
  #gives different result due to different calculation order
  # P = cpt:::change_point_test_pvalue_old(
  data,
  q_max = 6,
  n = n
)


# plot results (- log(P) vs k for q = 1, .. , q_max)

no.of.t <- NROW(P)
k <- c(1:no.of.t)
s <- seq(no.of.t - 1)

par(mfrow = c(3, 2))
for (i in seq_len(NCOL(P))) {
  plot(
    k,
    -log(P[k, i]),
    xlim = c(0, no.of.t + 1),
    ylim = c(0, log(n)),
    pch = 4,
    main = sprintf("q = %i", i)
  )
  y <- -log(P[k, i])
  segments(k[s], y[s], k[s + 1], y[s + 1])
  abline(h = -log(0.10), col = "blue")
  abline(h = -log(0.05), col = "red")
  abline(h = -log(0.01), col = "green")
}

```

## Fig 9

```{r, out.width="100%"}
fig9_data <- calculate_n_cp_by_q(tf7, alpha = 0.05)
ggplot(data = fig9_data, mapping = aes(x = q, y = n_cp, group = track_id, colour = track_id)) +
  geom_smooth(span = .4) +
  geom_point()
```

